Thesis: draft

### 2021-05-19 draft.tex
incorporated feedback ch. 1 and ch.2

\addcontentsline{toc}{chapter}{Abstract}
\noindent
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
\newline
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downsides in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the appearance and formation of phenomena such as polarization, spread of extremism, formation of echo chambers, etc.\\
\newline
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
\newline
In this thesis toy models are designed to investigate the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions. We will investigate whether network properties such as community structure and clustering coefficient may enhance or hamper the formation of echo chambers. We will use some basic filtering algorithms to determine their possible effects on the appearance of echo chambers and we will investigate the impact of individual stubbornness. Beside that, we will also pay some attention to the possible ways of distributing the opinions and how this may affect the opinion dynamics in the system. \\
\newline
Finally, we will test the model on real-world networks. Some question that may arise are: Do the results agree with those obtained from the theoretical network models? What are the similarities or differences in the results? What are the possible underlying reason/structures for these similarities/differences?
% definetely not finished!! --> add some type of conclusions or not? Change part about distributing stubborn people??
\setcounter{abstractpage}{\value{page}}
\end{abstract}

@@

Het is gekend dat mensen niet enkel hun mening uit zichzelf vormen, maar sterk beïnvloed worden door onder andere de mensen rond hen, hun omgeving, hun opvoeding,... Sociale media laten onmiddellijke interacties toe tussen mensen van overal op de wereld, zonder daarbij belemmerd te worden door fysieke en/of tijdsgrenzen. Dit kan een potentieel grote impact hebben op hoe meningen zich vormen en hoe ze evolueren.\\

@@

Door het toenemende gebruik van sociale media worden deze echter overspoeld door informatie. Om hun gebruikers niet het gevoel te geven dat ze verdrinken in wat ze te zien krijgen op hun sociale media platform, gebruiken deze sociale media platformen filtering algoritmen. Deze filtering algoritmen filteren, zoals de naam al impliceert, en individualiseren de informatie die iemand te zien krijgt op zijn/haar tijdslijn. Een gebruiker op sociale media krijgt dus niet alle berichten die zijn/haar vrienden posten te zien, maar enkel een selectie die onder andere gemaakt wordt op basis van de gebruiker zijn/haar sociale media geschiedenis. Dit filteren/individualiseren kan, mogelijks, grote gevolgen hebben. Zo zou het kunnen leiden tot de vorming van ``echo kamers'' waar gebruikers gevangen raken tussen ideeën die hun eigen visie versterken. Dit kan op zijn beurt polarisatie in de maatschappij, opkomst van extremisme, enzovoort bevorderen. Het is daarom uiterst belangrijk om de impact van deze filtering algoritmen grondig in kaart te brengen.\\

@@

Mensen zijn echter ook in staat om kritisch te denken, soms ook over-kritisch of selectief kritisch, maar dat terzijde. Dit kritisch denken kan ertoe leiden dat mensen niet zomaar hun mening veranderen omdat hun vrienden en/of omgeving dat suggereren. Ze zijn meer overtuigd van hun eigen gelijk of menen dat hun eigen visie een correcter standpunt inneemt. Deze individuele koppigheid zou men niet mogen verwaarlozen in het ontwerpen van modellen die de dynamica van opinies op sociale media willen onderzoeken.\\

@@

In deze thesis wordt gebruik gemaakt van ``toy models'' die de interactie tussen verschillende netwerk modellen, filtering algoritmen en individuele terughoudenheid om van mening te veranderen onderzoeken in de evolutie van twee opinies. Er worden verschillende netwerk modellen geïntroduceerd die de impact van onder andere clustering en/of groepsstructuur op de evolutie van opinies en de vorming van echo kamers moeten onderzoeken. Daarnaast wordt er gebruik gemaakt van enkel simpele filtering algoritmen om hun impact op de dynamica van de opinies te onderzoeken. Tot slot wordt er een notie van koppigheid in het systeem geïntroduceerd om mogelijke effecten hiervan in kaart te kunnen brengen. \\

@@

Naast het bovenstaande zal er ook kort aandacht gespendeerd worden aan hoe de twee verschillende opinies verdeeld worden over het systeem. Is er bijvoorbeeld een verschil tussen de twee meningen willekeurig over het systeem te verspreiden versus deze in een meer groepsstructuur te verdelen? Om de thesis te besluiten zal het opinie-dynamica model, dat in de thesis gebruikt werd, getest worden op netwerken verkregen van data uit de werkelijke wereld (bv. van email data of telefoonnetwerken). Er zal nagegaan worden of de resultaten van deze werkelijke netwerken gelijkenissen, dan wel verschillen, vertonen met de data verkregen van de netwerk modellen.

@@

% ADD MORE REFERENCES (previous research, contextualization etc)

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. It is important to understand the effects of individual resistance on the formation and evolution of opinions. When designing models of opinion dynamics on social networks, this individual stubbornness should not be ignored.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\
\newline
Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

\section{Complex networks}\label{complNet}

Complex networks have become more and more popular for analyzing complex, dynamical systems. They are used in fields such as physics, economics, social sciences, biology, etc \cite{Costa2008}. These different fields are very diverge, but they have at least one common ground: they often deal with a large number of variables/components that interact with each other. In other words, in all these fields one encounters complex systems, systems where it is not possible to predict collective behavior based on the properties of the individual components alone \cite{Mata2020}. Complex systems often display phenomena such as non-linearity, emergence, spontaneous order,...
One of the tools to deal with these complex systems and to give us more insight in the possible underlying structures are complex networks. A network, also often called a graph (both terms will be used in this thesis), is a structure composed of nodes or vertices and a set of links or edges that indicate the interactions between the nodes (in this thesis terms like agent, actor, user, individual, people, etc will be interchangebly used to denote the nodes; edges will usually just be called edges) \cite{Mata2020}.\\
Representing/modeling a complex system as a graph makes the system appear more simple and tractable, while it still includes the non-linearity of these systems. One can find the language to describe networks in mathematical graph theory. However, complex systems in real-life situations often deal with a huge number of components, so the use of statistical and high-performance computing tools is inevitable \cite{Mata2020}. One could argue that the study of complex networks lies somewhere at the intersection of mathematical graph theory and statistical physics \cite{F.Costa2007}.\\
The advantage of working with network models is that they reduce the level of complexity encountered in the real world, so that one can treat these systems in a more practical way. However, we still want our models to display properties similar to the ones seen in real systems \cite{Mata2020}. Since many real systems are not static but evolve in time, this means that we do not only need to deal with static networks but also with temporal networks. Temporal networks are networks where the edges are not always active, but instead become active for some periods of time \cite{Holme2012}. In temporal networks time becomes an explicit element in the network representation \cite{Holme2012}. Static networks have been widely studied and are often convenient for their analytic tractability, whereas temporal networks are, in some cases, more realistic \cite{Mata2020}.\\
Some other important concepts in network theory are the degree distribution, clustering/community structure, connectivity, etc. These concepts will be explained in depth in Chapter \ref{chap2}, Section \ref{netDef}, but let us already anticipate a bit on the case of social networks. It is found that many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale-free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering. These two are also referred to as ``small world phenomenon'' (note that in the literature the term ``small world'' sometimes only refers to the property of short distances, however, in this thesis, we will use it to indicate both a high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of them. However, they might still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\
Some examples of theoretical network models used in this thesis are the Erd\H{o}s-R\'{e}nyi network, the stochastic block model and the Watts-Strogatz model. These will be explained thoroughly in Chapter \ref{chap2}, Section \ref{netModel}.

% talk about difference between simple networks and complex network??

% talk about using real-world networks to compare against network models??

@@

The formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc. Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
\newline
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in this thesis will be given.

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. The use of statistical physics may, however, also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}\cite{Gug2015}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality \cite{Gug2015}. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in regular physical research and science in general.

% talk somewhere about rise of social physics, the need to use statistical physics (because of macroscopic phenomena as scale invariance, emergence, polarization,...) --> partly done?

@@

This thesis will try to investigate the interplay between the social network structure and the individual resistance to change in the evolution of two competing opinions. On one hand, we want to determine the influence of different network structures on the formation and evolution of opinions, whereas, on the other hand, we would like to investigate the effect of stubborn actors in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms must also be included.\\ 
\newline
In one of their previous papers (\textit{``Modelling opinion dynamics in the age of algorithmic personalization''}, \cite{Perra2019}) Perra and Rocha investigated both the effect of different filtering algorithms and different network topologies on the formation and evolution of two competing opinions. They also considered the effect of nudging (meaning that one opinion is pushed to all agents in the network). Their main findings were that the algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering coefficient) could result in the formation of echo chambers \cite{Perra2019}. On the other hand, a more heterogeneous contact pattern hampers this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (namely semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging \cite{Perra2019}. They, however, did not investigate the effect of stubborn actors in the network, nor did they investigate networks with a community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have a community structure and relatively low clustering to networks without community structure, but with a high clustering coefficient (a detailed description of concepts like clustering coefficient and community structure will be given in Chapter \ref{chap2}, Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with a high clustering coefficient. We will also pay some attention to networks that have both a high clustering coefficient and community structure. Can we observe echo chambers in those kinds of networks? If yes, can we determine which property might be the main driver for the formation of echo chambers: high clustering or community structure? \\
\newline
On the other hand, we want to determine the impact of stubborn actors on the formation and evolution of opinions in the network. Some interesting questions that may come to mind are: Is there an interplay between network structure and stubbornness? What about the possible interplay between the filtering algorithms and the introduction of stubborn actors?\\
% partly rewrite this according to results? Convergence to one opinion in communities --> discussed? Clustered stubborn people --> discussed? Include clustered opinions? Include real-world networks?

@@

A triangle is defined as a loop of length three, this is a sequence of nodes $x, y, z, x$ such that $\{x, y\}, \{y, z\}$ and $\{z, x\}$ are edges of the network. The clustering coefficient is thus a way to measure the degree to which nodes in a network tend to cluster \cite{Li2017}. The local clustering coefficient of a node $i$ is defined as \cite{Mata2020}

@@

	cc_i = \frac{2n_i}{k_i(k_i-1)} \ ,

@@

where $n_i$ is the number of edges that actually exist between the nodes in the neighborhood of $i$ and $k_i(k_i-1)/2$ is the maximum number of edges that could exist between them (this expression is only valid for an undirected graph, since for a directed graph $e_{ij} \neq e_{ji}$ and we have $k_i(k_i-1)$ possible edges between the neighbors of node i). % should give more mathimatical expression for n_i?
The average clustering coefficient is then given as the average of the local clustering coefficients \cite{Newman2003}

@@

\textbf{Including a stubbornness parameter}\\

>>>>

\addcontentsline{toc}{chapter}{Abstract}
\noindent
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
\newline
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downsides in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the formation of phenomena such as polarization, spread of extremism and formation of echo chambers.\\
\newline
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
\newline
In this thesis toy models are designed to study the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions. We will investigate whether network properties such as community structure and clustering may enhance or hamper the formation of echo chambers. We will use filtering algorithms to determine their possible effects on the formation of echo chambers and we will study the impact of individual stubbornness. We will also study different ways of distributing the opinions and how this may affect the opinion dynamics in the system. \\
\newline
Finally, we will test the model on real-world networks. We will investigate whether the results agree with those obtained from the theoretical network models.

\setcounter{abstractpage}{\value{page}}
\end{abstract}

@@

Het is gekend dat mensen niet enkel hun mening uit zichzelf vormen, maar sterk beïnvloed worden door onder andere de mensen rond hen, hun omgeving, hun opvoeding. Sociale media laten onmiddellijke interacties toe tussen mensen van overal op de wereld, zonder daarbij belemmerd te worden door fysieke en/of tijdsgrenzen. Dit kan een potentieel grote impact hebben op hoe meningen zich vormen en hoe ze evolueren.\\

@@

Door het toenemende gebruik van sociale media worden deze echter overspoeld door informatie. Om hun gebruikers niet het gevoel te geven dat ze verdrinken in wat ze te zien krijgen op hun sociale media platform, gebruiken deze sociale media platformen filtering algoritmen. Deze filtering algoritmen filteren, zoals de naam al impliceert, en individualiseren de informatie die iemand te zien krijgt op zijn/haar tijdslijn. Een gebruiker op sociale media krijgt dus niet alle berichten die zijn/haar vrienden posten te zien, maar enkel een selectie die onder andere gemaakt wordt op basis van de gebruiker zijn/haar sociale media geschiedenis. Dit filteren en individualiseren kan, mogelijks, grote gevolgen hebben. Zo zou het kunnen leiden tot de vorming van ``echo kamers'' waar gebruikers gevangen raken tussen ideeën die hun eigen visie versterken. Dit kan op zijn beurt polarisatie in de maatschappij, opkomst van extremisme, enzovoort bevorderen. Het is daarom uiterst belangrijk om de impact van deze filtering algoritmen grondig in kaart te brengen.\\

@@

Mensen zijn echter ook in staat om kritisch te denken, soms ook over-kritisch of selectief kritisch, maar dat terzijde. Dit kritisch denken kan ertoe leiden dat mensen niet zomaar hun mening veranderen omdat hun vrienden of omgeving dat suggereren. Ze zijn meer overtuigd van hun eigen gelijk of menen dat hun eigen visie een correcter standpunt inneemt. Deze individuele koppigheid zou men niet mogen verwaarlozen in het ontwerpen van modellen die de dynamica van opinies op sociale media willen onderzoeken.\\

@@

In deze thesis wordt gebruik gemaakt van ``toy models'' die de interactie tussen verschillende netwerk modellen, filtering algoritmen en individuele terughoudenheid om van mening te veranderen onderzoeken in de evolutie van twee opinies. Er worden verschillende netwerk modellen geïntroduceerd die de impact van onder andere clustering en groepsstructuur op de evolutie van opinies en de vorming van echo kamers moeten onderzoeken. Daarnaast wordt er gebruik gemaakt van enkel simpele filtering algoritmen om hun impact op de dynamica van de opinies te onderzoeken. Tot slot wordt er een notie van koppigheid in het systeem geïntroduceerd om mogelijke effecten hiervan in kaart te kunnen brengen. \\

@@

Naast het bovenstaande zal er ook kort aandacht gespendeerd worden aan hoe de twee verschillende opinies verdeeld worden over het systeem. We zullen de opinies in groepsstructuren verdelen en dit scenario vergelijken met het scenario waarin de opinies willekeurig verdeeld worden. Om de thesis te besluiten zal het opinie-dynamica model, dat in de thesis gebruikt werd, getest worden op netwerken verkregen van data uit de werkelijke wereld (bv. van email data of telefoonnetwerken). Er zal nagegaan worden of de resultaten van deze werkelijke netwerken gelijkenissen, dan wel verschillen, vertonen met de data verkregen van de netwerk modellen.

@@

% ADD MORE REFERENCES (previous research, contextualization)

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have an impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. It is important to understand the effects of individual resistance on the formation and evolution of opinions. When designing models of opinion dynamics on social networks, this individual stubbornness should not be ignored.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant and convenient experience on their platforms, social media companies use algorithms to order and filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate models to reproduce the opinion dynamics and opinion formation processes found in real life.\\
\newline
Section \ref{statPhys} introduces concepts such as statistical and social physics. Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics, is introduced in Section \ref{Opinion}. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

An important tool in the study of opinion dynamics in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms and molecules to macroscopic observed behavior and has a wide range of applications inside and outside physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics \cite{Castellano2009}\cite{Galam2008}\cite{Galam1982}\cite{Stauffer2012}. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display regularities, transitions from disorder to order, the emergence of consensus, universality \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the regularities at large scale as collective effects of the interactions among individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one difficulty compared to the description of gases and other physical phenomena. In physics, the microscopic constituents are atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics of individuals would be well know, the interactions between them would still be complicated and it would be hard to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, extracting useful and qualitative information out of social models is a challenging task \cite{Castellano2009}. The use of statistical physics may, however, offer a way out. In statistical physics it is found that different systems behave qualitatively the same and are governed by the same laws near the critical point \cite{Hu2018}\cite{Kadanoff2010}. This lead to the concept of universality: many different microscopic problems have the same macroscopic manifestation \cite{Gug2015}\cite{Kadanoff2015}. This idea may then be extended to social systems \cite{Hu2018}: we should not take the unique character of every individual into account, instead these individual characteristics can be incorporated in a sort of averaged way \cite{Gug2015}. One can then proceed to build models of social behavior where only the most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with empirical data to investigate whether the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in physical research and science in general.
\newpage
\noindent
\section{Complex networks}\label{complNet}

Complex networks have become more and more popular for analyzing complex, dynamical systems. They are used in fields such as physics, economics, social sciences and biology \cite{Costa2008}. These different fields are diverse, but they have at least one common ground: they often deal with a large number of components that interact with each other. In other words, in all these fields one encounters complex systems, systems where it is not possible to predict collective behavior based on the properties of the individual components alone \cite{Mata2020}. Complex systems often display phenomena such as non-linearity, emergence, spontaneous order.
\\
\newline
One of the tools to deal with these complex systems and to give us more insight in the possible underlying structures are complex networks. A network is a structure composed of nodes or vertices and a set of links or edges that indicate the interactions between the nodes (in this thesis nodes will mostly be called actors; edges will be called edges) \cite{Mata2020}.\\
\newline
Representing a complex system as a network makes the system appear simpler and tractable, while it still includes the non-linearity of these systems. One can find the language to describe networks in mathematical graph theory. However, complex systems in real-life often deal with a large number of components, so the use of statistical and high-performance computing tools is inevitable \cite{Mata2020}. One could argue that the study of complex networks lies somewhere at the intersection of mathematical graph theory and statistical physics \cite{F.Costa2007}.\\
\newline
The advantage of working with network models is that they reduce the level of complexity encountered in the real world, so that one can treat these systems in a more practical way. However, we still want our models to display properties similar to the ones seen in real systems \cite{Mata2020}. Since many real systems are not static but evolve in time, this means that we do not only need to deal with static networks but also with temporal networks. Temporal networks are networks where the edges are not always active, but instead become active for some periods of time \cite{Holme2012}. In temporal networks time becomes an explicit element in the network representation \cite{Holme2012}. Static networks have been widely studied and are often convenient for their analytic tractability, whereas temporal networks are, in some cases, more realistic \cite{Mata2020}.\\
\newline
Some other important concepts in network theory are the degree distribution, clustering, community structure and connectivity. These concepts will be explained in depth in Chapter \ref{chap2}, Section \ref{netDef}, but let us already anticipate on the case of social networks. It is found that many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale-free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering, also referred to as ``small world phenomenon'' (note that in the literature the term ``small world'' sometimes only refers to the property of short distances, however, in this thesis, we will use it to indicate both a high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of them. However, they might still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\
\newline
The theoretical network models used in this thesis are the Erd\H{o}s-R\'{e}nyi network, the stochastic block model and the Watts-Strogatz model. These will be explained thoroughly in Chapter \ref{chap2}, Section \ref{netModel}.

@@

The formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (e.g. stubbornness) and the information someone is exposed to. Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or actors in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization and formation of echo chambers (clusters of people with the same opinion) \cite{Sirbu2016}. \\
\newline
The opinion of the actors in the model can be either discrete or continuous. This thesis deals with models where each actor can have one of the two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (e.g. republicans or democrats, renting or buying a house). \\
\newline
The rules that determine how an actor updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a probabilistic way of updating. An overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
\newline
One can also include a resistance parameter that determines the hesitation of an actor to change to a new opinion or a parameter that determines the influence of an actor on others. Furthermore, the actors can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of actors is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define different filtering algorithms. The filtering algorithms that social media companies use, are corporate secrets, but they typically have three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in this thesis will be given.

@@

This thesis will study the interplay between the social network structure and the individual resistance to change in the evolution of two competing opinions. On one hand, we will determine the influence of different network structures on the formation and evolution of opinions, whereas, on the other hand, we will study the effect of stubborn actors in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms will be included.\\ 
\newline
Perra and Rocha investigated both the effect of different filtering algorithms and different network topologies on the formation and evolution of two competing opinions \cite{Perra2019}. They also considered the effect of nudging (meaning that one opinion is pushed to all actors in the network). Their main findings were that the algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering) could result in the formation of echo chambers. On the other hand, a more heterogeneous contact pattern hampers this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging \cite{Perra2019}. They, however, did not study the effect of stubborn actors in the network, nor did they investigate networks with community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have community structure and relatively low clustering to networks without community structure, but with high clustering (a detailed description of concepts like clustering and community structure will be given in Chapter \ref{chap2}, Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with high clustering. We will also study networks that have both high clustering and community structure to identify the conditions and driving forces leading to the formation of echo chambers.\\
\newline
We will also study the impact of stubborn actors on the formation and evolution of opinions in the network. We aim to determine whether there is an interplay between network structure, filtering algorithm and stubbornness.

@@

A triangle is defined as a loop of length three, this is a sequence of nodes $x, y, z, x$ such that $\{x, y\}, \{y, z\}$ and $\{z, x\}$ are edges of the network. The clustering coefficient is thus a way to measure the degree to which nodes in a network tend to cluster \cite{Li2017}. The local clustering coefficient of a node $i$ is defined as 

@@

	cc_i = \frac{2e_i}{k_i(k_i-1)} \ ,

@@

where $e_i$ is the number of edges that actually exist between the nodes in the neighborhood of $i$ and $k_i(k_i-1)/2$ is the maximum number of edges that could exist between them \cite{Mata2020} (this expression is only valid for an undirected network, since for a directed network $e_{ij} \neq e_{ji}$ (where $e_{ij}$ denotes an edge from node $i$ to node $j$) and we have $k_i(k_i-1)$ possible edges between the neighbors of node i). % should give more mathimatical expression for n_i?
The average clustering coefficient is 

@@

\textbf{Including a resistance parameter}\\
